---
title: "R Notebook"
output:
  html_document:
    df_print: paged
editor_options: 
  markdown: 
    wrap: 72
---

## Problem

### Business Tasks

Cyclistic is a fictional bike-share company inside Chicago. This company
tasked us with the analysis of this question:\
How do annual members and casual riders **use** Cyclistic bikes
differently?\
The *"casual" riders* are the clients that use single-ride and full day
pass. Those offer are about **short term engagement.**\
The *"member" riders* are the clients that buy annual membership. This
offer is about **long term engagement**.\
In order to convert "casual" clients into "member" clients, we could try
to develop **new "transition" offer** that is about *mid term
engagement*. This offer would be developed to be **attractive to
"casual" clients** while being **unattractive to "member"** clients
since we don't want to convert members into casuals. In order to find
opportunities for said new transition offers, we will analyse the riders
behaviors.\
Let's break this business task into sub tasks:

-   Identify and categorize how users use Cyclistic services
-   See how those usage relate to casual riders and annual members
-   Identify usage differences and similarities between casual riders
    and annual members

### Audience (stakeholders in **bold**):

-   **Director of marketing**
-   Part of analytics team
-   **Cyclistic executives**

### Data sources:

-   [Cyclistic's first-party
    data](https://divvy-tripdata.s3.amazonaws.com/index.html) from the
    last 12 months: from march 22 to February 23. Cyclistic is a
    fictional company so the data is provided by Motivate International
    Inc. The data contain one row per user's trip. The variables are
    about:

    -   trip's starting hour and location
    -   trip's ending hour and location
    -   customer's membership
    -   bike type

    The files used are:

    -   2022MM-divvy-tripdata/2022MM-divvy-tripdata.csv with MM the
        month (03 to 12)
    -   202209-divvy-tripdata/202209-divvy-publictripdata.csv
    -   2022MM-divvy-tripdata/2022MM-divvy-tripdata.csv with MM the
        month (01 to 02)

-   [Chicago ZIP
    boundaries](https://data.cityofchicago.org/Facilities-Geographic-Boundaries/Boundaries-ZIP-Codes/gdcf-axmw)
    Contains a shapefile map that represents the Chicago divided by ZIP
    areas

```{r setup}
dataPath <- 'absolutePath'
knitr::opts_knit$set(root.dir = normalizePath(dataPath)) 
```

## Analysis

```{r include=FALSE}
# install.packages("readr")
# install.packages("skimr")
#install.packages("tidyverse")
#install.packages("data.table")
# install.packages("sf")
# library("readr")
library("skimr")
library(tidyverse)
library(data.table)
library(sf)
library(lubridate)
library(glue)
# testData <- read.csv("./202302-divvy-tripdata/202302-divvy-tripdata.csv")
# skim_without_charts(testData)
```

```{r include=FALSE}
isTripFile <- function(filePath){
  readLines(filePath, n=1) %>% 
    grepl(pattern = "ride_id|trip_id") 
}
# testFileName <- "enterFileName"
# isTripFile(glue("{dataPath}/{testFileName}"))

dataFileNames <- list.files(recursive = TRUE, pattern = "*.csv") %>%
  str_split(string = ., pattern = "\n") 
dataFileNames <- dataFileNames[mapply(isTripFile, dataFileNames)]

tripDataFiles <- dataFileNames %>%
  grep(x = dataFileNames, pattern = '20(2[2-3])', value = TRUE) %>% #Select years in range 2020-2023
  grep(x = ., pattern = '20220[1-2]', invert = T, value = T) %>% 
  grep(pattern = 'station',  x = ., ignore.case = TRUE, invert = TRUE, value = TRUE) %>% #Remove files on stations
  lapply(fread) %>% #Read data from the selected file
  rbindlist() #Merge data from selected file

tripData <- tripDataFiles
rm(tripDataFiles)
```

```{=html}
<!--Changes in data model accross the years (not very useful since we only need to study the last year):

* from 01 - 2020 : ride_id,rideable_type,started_at,ended_at,start_station_name,start_station_id,end_station_name,end_station_id,start_lat,start_lng,end_lat,end_lng,member_casual
!!! column names surrounded by quotes from 09 - 2022
Change: 
trip_id -> ride_id
start_time -> started_at
end_time -> ended_at
bikeid -> 0
tripduration -> 0
from_station_id -> start_station_id
from_station_name -> start_station_name
to_station_id -> end_station_id
to_station_name -> end_station_name
usertype -> member_casual
gender -> 0
birthyear -> 0
* from 01 - 2014 : trip_id,start_time,end_time,bikeid,tripduration,from_station_id,from_station_name,to_station_id,to_station_name,usertype,gender,birthyear
!!! Q2-2019, Q1-2018, with quote around column names sometimes
Change: starttime, stoptime, birthday -> start_time, stop_time, birthyear
* from 01 - 2013 : trip_id,starttime,stoptime,bikeid,tripduration,from_station_id,from_station_name,to_station_id,to_station_name,usertype,gender,birthday
-->
```
```{r eval=FALSE, include=FALSE}
#Unused code sketch to load and convert data from previous years
#THIS CODE IS DISABLED AND WILL BUG IF ABLED

dateTimeFormat <- c("%m/%d/%Y %H:%M:%S", "%m/%d/%Y %H:%M")

tripDataFiles15_16 <- dataFileNames %>%
  grep(x = dataFileNames, pattern = '20(1[5-6])', value = TRUE) %>% #Select years in range 2015-2016
  grep(pattern = 'station',  x = ., ignore.case = TRUE, invert = TRUE, value = TRUE) %>% #Remove files on stations
  lapply(fread) %>% #Read data from the selected file
  rbindlist() %>%  #Merge data from selected file
  rename(
    started_at = starttime,
    ended_at = stoptime
  ) %>%
  mutate(started_at = parse_date_time(started_at,dateTimeFormat), ended_at = parse_date_time(ended_at,dateTimeFormat)) #Convert datetime string to datetime

tripDataFiles17 <- dataFileNames %>%
  grep(x = dataFileNames, pattern = '2017', value = TRUE) %>% #Select years 2017
  grep(pattern = 'station',  x = ., ignore.case = TRUE, invert = TRUE, value = TRUE) %>% #Remove files on stations
  lapply(fread) %>% #Read data from the selected file
  rbindlist() %>%  #Merge data from selected file
  rename(
    started_at = start_time,
    ended_at = end_time
  ) %>%
  mutate(started_at = parse_date_time(started_at,dateTimeFormat), ended_at = parse_date_time(ended_at,dateTimeFormat)) #Convert datetime string to datetime

tripDataFiles18_19 <- dataFileNames %>%
  grep(x = dataFileNames, pattern = '20(1[8-9])', value = TRUE) %>% #Select years in range 2014-2019
  grep(pattern = 'station',  x = ., ignore.case = TRUE, invert = TRUE, value = TRUE) %>% #Remove files on stations
  lapply(fread) %>% #Read data from the selected file
  rbindlist() %>%  #Merge data from selected file
  rename(
    started_at = start_time,
    ended_at = end_time
  )

tripDataFiles <- rbindlist( list(tripDataFiles15_16, tripDataFiles17)) %>%
  rename(
    ride_id = trip_id,
    start_station_id = from_station_id,
    start_station_name = from_station_name,
    end_station_id = to_station_id,
    end_station_name = to_station_name,
    member_casual = usertype
  )
```

```{r include=FALSE}
# Read in shapefile of Chicago map
chicago_sf <- st_read(dsn = list.files('./ChicagoBoundariesZIPCodes', full.names = T, pattern = "*.shp")) %>% 
  st_transform(., 4326)

# Convert to data frame for use with ggplot
chicago_df <- fortify(chicago_sf) %>% 
  st_transform(crs = 4326)
# rm(chicago_sf)

coordinatesPrecision<- 3 #With 3 digits the coordinate precision is around 110m
chicagoBbox<-st_bbox(chicago_sf)

#List all the coordinates in Chicago up to a selected precision
chicagoCoordinates <- expand.grid(seq(chicagoBbox$xmin, chicagoBbox$xmax, 0.1^coordinatesPrecision), seq(chicagoBbox$ymin, chicagoBbox$ymax, 0.1^coordinatesPrecision) ) %>% #Cartesian product of chicago lat and long arrays
  round(. ,coordinatesPrecision) %>%
  setnames(c("lng","lat")) %>% 
  mutate(lt = lat, lg= lng) %>% 
  st_as_sf(coords = c("lg","lt"), crs = 4326)

#Associate every coordinates in chicagoCoordinate to a zone
coordinateToZoneTable <- st_join(x=chicagoCoordinates, y=chicago_df, join = st_within, left=T) %>% #Coordinate to zone mapping
  drop_na() %>%
  as.data.frame() %>% 
  select(c(lat, lng, objectid))

rm(chicago_sf, chicagoCoordinates, chicagoBbox)
```

### Data cleaning

#### Removed data:

-   **duplicate** trips
-   the trips that **last for a minute or less** because it's a canceled
    trip
-   the trips that **last for more than a day** because it's likely that
    the user forgot to return the bike
-   the trips that had **starting or ending location outside of
    Chicago** because that's not supposed to be possible
-   the trips with a duration that are in the **1% longest trips** to
    make readable the graphs that have tripduration as an axis

#### Data modifications:

-   added **trip duration**
-   **deteriorated the coordinates** so that the possible number of
    coordinates inside Chicago is finite
-   added **day** of the week
-   added **month**
-   added ZIP **zone IDs** for the trips *starting location* and *ending
    location*
-   **ordered** by *station name* and then by *start time*

```{r include=FALSE}
processedData <- tripData %>%
  drop_na(end_lat, end_lng) %>% #Drop trajects without end coordinates
  unique() %>% #Remove duplicates
  mutate(tripduration = difftime(ended_at, started_at)) %>%   #Add trips duration
  filter(tripduration>=60) %>% #Filter trip of less than 1min
  filter(tripduration<=86400)  #Filter trip of more than 1day
  
processedData <- processedData %>% 
  filter(tripduration<quantile(processedData$tripduration, probs = c(.99))) %>%   #Filter outliers
  mutate(day = (wday(started_at)-2)%%7 +1) %>% #Add day of the week
  mutate(month = month(started_at)) %>% #Add month of the year
  mutate(start_lat = round(start_lat, coordinatesPrecision) , start_lng = round(start_lng, coordinatesPrecision) , end_lat = round(end_lat, coordinatesPrecision) , end_lng = round(end_lng, coordinatesPrecision) ) %>%  #Deteriorate lat, long precision to have less distinct value
  merge(., coordinateToZoneTable, by.x=c("start_lat", "start_lng"), by.y = c("lat", "lng")) %>% #Add the zone id where the trip started. This also filter the trip that started outside of Chicago
  setnames(., "objectid", "startZoneId") %>% 
  merge(., coordinateToZoneTable, by.x=c("end_lat", "end_lng"), by.y = c("lat", "lng")) %>% #Add the zone id where the trip ended. This also filter the trip that ended outside of Chicago
  setnames(., "objectid", "endZoneId") %>%  
  setorderv(c("start_station_name","started_at"), c(1,1)) #Order data by station then start time
  
rm(tripData, coordinateToZoneTable)
```

### Usage by time

#### Usage duration

Both casual and member clients trip's duration is around **5 min** but
members are more concentrated around that duration.

```{r echo=FALSE, message=FALSE, warning=FALSE}
durationHisto <- hist(as.numeric(processedData$tripduration), plot = F)
maxFreq <- which.max(durationHisto$counts)
maxFreqValue <- mean(durationHisto$breaks[c(maxFreq, maxFreq+1)])[1]

ggplot(processedData[member_casual=="casual"], aes(x=tripduration, y =  after_stat(count / sum(count))))+
  geom_histogram(aes(fill='casual'), alpha = 0.5)+
  geom_histogram(data=processedData[member_casual=="member"], aes(fill='member'), alpha = 0.5)+
  scale_y_continuous(labels = scales::percent)+
  scale_fill_discrete(name = "Customer types")+
  labs(x = "Trip duration", y = "Percentage of subset trips", title = "Trip duration by customer type", subtitle = "Trips subsets are based on customer type. The shown percentages are relative to the subset")+
  geom_vline(xintercept = maxFreqValue)+
  annotate("text", x = maxFreqValue, y = 0.01, label = as.character(maxFreqValue))

rm(durationHisto, maxFreq, maxFreqValue)
```

#### Usage throughout the day

Casuals use the service all along the day, with a cliff around 5:30 AM
and a **peak around 5:30 PM**\
Member's graph is similar casual's graph with two variations:

-   another **peak around 8:00 AM**
-   less usage during working hours

```{r echo=FALSE, message=FALSE, warning=FALSE}
timeData<-processedData[,c("started_at", "member_casual")] %>% 
  mutate(time= ymd_hms(paste0(ymd("2000-01-01"), format(started_at, "%H:%M:%S"))) ) #We set all the observations to the same day

morningCommuteHour<-ymd_hms("2000-01-01 08:00:00")
eveningCommuteHour<-ymd_hms("2000-01-01 17:30:00")

ggplot(timeData[member_casual=="casual"], aes(x=time, y =  stat(count / sum(count))) )+
  geom_histogram(aes(fill='casual'), alpha = 0.5)+
  geom_histogram(data=timeData[member_casual=="member"], aes(fill='member'), alpha = 0.5)+
  scale_y_continuous(labels = scales::percent)+
  scale_x_datetime(date_labels = "%H:%M", date_breaks = "2 hour")+
  scale_fill_discrete(name = "Customer type")+
  labs(x = "Hour of day", y = "Percentage of subset trips", title = "Loaning hour by customer type", subtitle = "Trips subsets are based on customer type. The shown percentages are relative to the subset")+
  geom_vline(xintercept = morningCommuteHour)+
  annotate("text", x = morningCommuteHour, y = 0.09, label = "Morning commute hour")+
  geom_vline(xintercept = eveningCommuteHour)+
  annotate("text", x = eveningCommuteHour, y = 0.09, label = "Evening commute hour")

rm(timeData, morningCommuteHour, eveningCommuteHour)
```

#### Usage throughout the week

*Casuals* use Cyclistic a lot more during the **weekend** than during
the workweek. On the opposite, *members* have a close to **constant
usage during the whole week** but they use the service slightly more
during the workweek than during the weekend.

```{r echo=FALSE}
day_labels <- c("Mon", "Tue", "Wed", "Thu", "Fri", "Sat", "Sun")

ggplot(processedData[member_casual=="casual"], aes(x = day, y = stat(count/sum(count)) )) +
  geom_bar(aes(fill='casual'), alpha = 0.5) +
  geom_bar(data = processedData[member_casual=="member"], aes(fill='member'), alpha = 0.5)+
  scale_fill_discrete(name = "Customer type")+
  labs(x = "Day of week", y = "Percentage of subset trips", title = "Loaning day by customer type", subtitle = "Trips subsets are based on customer type. The shown percentages are relative to the subset")+
  scale_x_continuous(breaks = seq(1, 7, 1), labels = day_labels)+
  scale_y_continuous(labels = scales::percent)+
  geom_vline(xintercept = 5.5)+
  annotate("text", x = 3, y = 0.21, label = "Workweek")+
  annotate("text", x = 6.5, y = 0.21, label = "Weekend")
```

#### Usage throughout the year

*Casuals* enjoy Cyclistic during the **sunny months** (more than 15%
usage during June to august) and avoid it during the cold ones (less
than 2.5% usage during the December to February).\
*Members* follow the same trend, but their usage is much more
**consistant**, with usage under 4% only in December and no month with
usage above 12.5%.

```{r echo=FALSE}
ggplot(processedData[member_casual=="casual"], aes(x = month, y = stat(count/sum(count)) )) +
  geom_bar(aes(fill='casual'), alpha = 0.5) +
  geom_bar(data = processedData[member_casual=="member"], aes(fill='member'), alpha = 0.5)+
  labs(x = "Months", y = "Percentage of subset trips", title = "Loaning months by customer type", subtitle = "Trips subsets are based on customer type. The shown percentages are relative to the subset")+
  scale_fill_discrete(name = "Customer type")+
  scale_x_continuous(breaks = seq(1, 12, 1), labels = month.abb)+
  scale_y_continuous(labels = scales::percent)+
  geom_vline(xintercept = 3.166)+
  geom_vline(xintercept = 6.166)+
  geom_vline(xintercept = 9.166)+
  geom_vline(xintercept = 12.166)+
  annotate("text", x = 1.666, y = 0.18, label = "Winter")+
  annotate("text", x = 4.666, y = 0.18, label = "Spring")+
  annotate("text", x = 7.666, y = 0.18, label = "Summer")+
  annotate("text", x = 10.666, y = 0.18, label = "Autumn")
```

### Social usage

We consider that, if the user begin their trip in the **same minute and
station** as another user, then he makes a 'social usage' of Cyclistic
services.\
This measure is **far from exact** because two user that don't know each
others can take a bike in the same minute and station. Or two user that
went on a bike trip could take more than a minute to lend the bike, or
they could start in different stations and meet while on bike.\
However uncertain, the results are interesting. It shows that *casuals*
use socially Cyclistic's services **35%** of the time while *members*
have a social usage only **20%** of the time.

```{r echo=FALSE}
socialData<-processedData[start_station_name!=""] %>% 
  mutate(social = (start_station_name==lag(start_station_name) & abs(difftime(started_at,lag(started_at)))<60) | (start_station_name==lag(start_station_name) & (abs(difftime(started_at,lead(started_at))))<60)) %>%  #Add a social column that is True if the user begins their trip in the same minute and station as another user
  drop_na()

socialMemberPercent = round(mean(socialData[member_casual=="member"]$social), 3)
socialCasualPercent = round(mean(socialData[member_casual=="casual"]$social), 3)

ggplot(socialData, aes(y = member_casual, fill = social)) +
  geom_bar(position = "fill")+
  scale_x_continuous(labels = scales::percent)+
  scale_fill_discrete(name = "Social Usage")+
  labs(x = "Percentage of social usage", y = "Customer type", title = "Social usage by customer type")+
  geom_vline(xintercept = socialMemberPercent)+
  annotate("text", x = socialMemberPercent, y = 2, label = glue("{socialMemberPercent*100}%"))+
  geom_vline(xintercept = socialCasualPercent)+
  annotate("text", x = socialCasualPercent, y = 1, label = glue("{socialCasualPercent*100}%"))

rm(socialData, socialMemberPercent, socialCasualPercent)
```

## Recommendations

The analysis aims at uncovering the differences between "casual" and
"members" clients in order to understand what could **convince a
"casual" client to become a "member" client**.\
To achieve this goal, we will make 3 recommendations that aims at
producing *transitional offer* for casuals toward annual pass or *making
the annual pass more interesting for casual*. Those recommendations are:

-   Creation of a **summer pass** since it's the season that interest
    casual the most. The member shall keep it's current offer to have
    access to the service during the whole year.
-   Creation of an **annual leisure pass** that gives:
    -   full access during weekends

    -   access between 19h-5h during workweek

        Those time ranges interest the casuals. The member shall keep
        his current offer to have access to the two commuting time
        range.
-   On annual passes, allow the member to **loan a 2nd bike for free**
    \~3 times per week. This could make long time offers more attractive
    to casuals and at the same times it is a way for non customers to
    test the service in very favorable conditions (social usage).

## Appendix

### Unused findings

#### Usage by location

##### Location alone

At first glance, both members and casuals make a similar usage of
Cyclistic in term of location alone: they are both centered on the
**Lincoln park** area of ZIP code 60614. However, by analyzing the
difference between member usage we can see that:

-   *casuals* prefer **Streeterville** neighborhood, **Lincoln park**
    and **South Loop** a bit more

-   *members* prefer **West Loop** and **Woodlawn** a bit more

These tendencies remain the same during the week, but get reinforced for
Streeterville during the weekend.\
We only display the graphs based on the trip ending zone because the
graph based on trip beginning zone are similar.

```{r echo=FALSE, message=FALSE, warning=FALSE}

options(dplyr.summarise.inform = FALSE)

#Function that plot:
#   - station usage in Chicago districts 
#   - station usage difference between user types in Chicago districts
plot_chicago_data_memberCasualDiff <- function(chicago_data, chicago_map, zoneType, contextualisation){
  customerTypeByZoneType <- chicago_data %>% 
    group_by(.data[[zoneType]], member_casual) %>%
    summarize(count = n()) %>% 
    pivot_wider(names_from = member_casual, values_from = count, values_fill = 0)
  
  chicago_map_with_customerTypeCount <- merge(chicago_map, customerTypeByZoneType, by.x=("objectid"), by.y=(zoneType))
  
  plots<-list()
    
  p<-ggplot(data = chicago_map_with_customerTypeCount) +
    geom_sf(aes(fill=((member+casual)/sum(member+casual))))+
    scale_fill_viridis_c(na.value = "pink",
                         labels = scales::percent_format(),
                         name = "% of usage")+
    theme(axis.text.x = element_text(angle = 45))+
    labs(title = glue("Chicago zone usage {contextualisation}")) 
  plots[[1]]<-p
  
  p<-ggplot(data = chicago_map_with_customerTypeCount) +
    geom_sf(aes(fill=( (member)/sum(member) - (casual)/sum(casual) )))+
    scale_fill_gradient2(low = "red",
                         mid = "lightgray",
                         high = "blue",
                         labels = scales::percent_format(),
                         name = "% of relative usage difference: member - casual")+
    theme(axis.text.x = element_text(angle = 45))+
    labs(title = glue("Chicago zone relative usage difference between users {contextualisation}"), subtitle = "The difference is calculated as member usage minus casual usage")  
  plots[[2]]<-p
  
  return(plots)
}

# processedData %>% 
#   plot_chicago_data_memberCasualDiff(., chicago_df, "startZoneId", "for trip beginning")

processedData %>% 
  plot_chicago_data_memberCasualDiff(., chicago_df, "endZoneId", "for trip end")
```

##### Location during workweek

Casuals are slightly more concentrated on Lincoln Park than members
during the workweek

```{r echo=FALSE, warning=FALSE}
# processedData %>% 
#   filter(day %in% c(1:5) ) %>% 
#   plot_chicago_data_memberCasualDiff(., chicago_df, "startZoneId", "for worweek trip beginning")

processedData %>%
  filter(day %in% c(1:5) ) %>%
  plot_chicago_data_memberCasualDiff(., chicago_df, "endZoneId", "for workweek trip end")
```

##### Location during weekend

Members are slightly more concentrated on Lincoln Park during the
weekend

```{r echo=FALSE, warning=FALSE}
# processedData %>% 
#   filter(day %in% c(6:7) ) %>%  
#   plot_chicago_data_memberCasualDiff(., chicago_df, "startZoneId", "for weekend trip beginning")

processedData %>%
  filter(day %in% c(6:7) ) %>% 
  plot_chicago_data_memberCasualDiff(., chicago_df, "endZoneId", "for weekend trip end")
```

#### Unrelated findings

```{r echo=FALSE}
processedData %>% 
  select(c(startZoneId, endZoneId)) %>% 
  pivot_longer(c(startZoneId, endZoneId), names_to = "zoneType", values_to = "zoneId") %>% 
  group_by(zoneId, zoneType) %>% 
  summarize(count = n()) %>% 
  pivot_wider(names_from = zoneType, values_from = count, values_fill = 0) %>% 
  merge(chicago_df, ., by.x=("objectid"), by.y=("zoneId")) %>% 
  ggplot(data = .) +
    geom_sf(aes(fill=( (endZoneId - startZoneId) )))+
    scale_fill_gradient2(low = "orange",
                         mid = "lightgray",
                         high = "green",
                         name = "Number of bike left in the zone")+
    theme(axis.text.x = element_text(angle = 45)) %>% 
  labs(title = "Chicago zone difference between bike take and bike leave over the year")  
```
